Evaluating infectious disease forecasts \\ with allocation scoring rules

### 2024-02-17 alloscore_manuscript/alloscore-application-shorter.Rnw
attempting to rework/excise ventilator narrative

We evaluated forecasts using the allocation score (AS) and the weighted interval score (WIS), computed at a horizon of
14 days. For this analysis, we fixed the resource constraint used for the AS to be $K=15,000$, based roughly on a
reported number of ventilators available for reallocation in the US \citep{ajao_assessing_2015}. We computed the mean
WIS (MWIS) across all of the forecasted state-level locations.

@@

cowplot::plot_grid(p1, p2, p4,
                   nrow=3,
                   align="v",
                   rel_heights = c(1, 1, 1.85),
                   labels = "AUTO")
@

@@

While our example used a hypothetical scenario to illustrate a potential application of the allocation score, resource
allocation decisions are a realistic example of the kinds of decisions that could motivate more targeted forecasting
exercises.
%One real-world example, which was the motivating example presented in the work above, is allocation of ventilators during a respiratory viral pandemic\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a
scarce resource during outbreaks \citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with
logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not
perfectly so) with hospitalization data. However, this example has limitations in that not everyone who is hospitalized
needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}. While
there may be limitations in this specific pairing of a resource and forecasted outcome, we emphasize that the important
result from this work is how evaluating forecasts using a fundamentally different kind of score can illuminate
previously hidden aspects of value in forecasts. In this setting, the feature of a forecast that emerges as being
important is that the probabilistic forecasts predict the relative magnitudes between locations accurately, even if the
absolute magnitudes are inaccurate.

In practice, there are many potential users of forecasts with many different decision making problems. While outcomes of
some decisions could potentially be specified by loss functions, other uses for forecasts may be less quantifiable, such
as when forecasts are used to inform a general sense of situational awareness that may feed in unmeasurable ways into
other decisions. Even those that are easily framed as expected loss minimization may differ enough that no single score
would be appropriate for all users. Ideally, targeted forecasting tools could be developed through close collaboration
between modelers and public health officials. However, this may only be possible in settings with sufficient staffing on
both an analytics and a public health team. Increasingly, collaborative modeling hubs are being used to generate
``one-size-fits-all'' forecasts for many locations at once. But it could still be beneficial in these settings to expand
the set of scoring methods used by the hub to include scores based on specific public health decision problems, such as
AS, allowing end users to better understand the value of available forecasts as inputs to their particular
decision making contexts. (We note that issues of propriety are raised when forecasts are evaluated with scoring rules
for which they were not elicited or via parameters imputed by a hub after submission, both of which are done in our
application example.  We consider these issues briefly in supplement sections \ref{sec:distfromq} and
\ref{sec:propriety_of_parametric_approximation}.)

>>>>

We evaluated forecasts using the allocation score (AS) and the weighted interval score (WIS), computed at a horizon of
14 days. We chose to focus much of our analysis on the AS computed for a resource constraint of $K=15,000$. This level provides
an anchor for interesting comparisons between phases of the Omicron wave by representing a national shortage at the peak and
a national excess before and after the wave (see Figure \ref{fig:metrics-over-time}A), assuming that the resource need corresponds 
directly to new hospital admissions.
We additionally computed the mean
WIS (MWIS) across all of the forecasted state-level locations.

@@

cowplot::plot_grid(p1, p2, p4,
                   nrow=3,
                   align="v",
                   rel_heights = c(1, 1, 1.85),
                   labels = "AUTO", hjust = -1)
@

@@

Our example application of the AS refers to a hypothetical and unspecified resource need that is supposed to 
pair with our forecasted outcome (new hospital admissions) in a manner that might appear overly simplistic and 
artificially direct.
A concrete decision problem, which motivated our work on allocation scoring, is

allocation of ventilators during a respiratory viral pandemic
\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a
scarce resource during outbreaks \citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with
logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not
perfectly so) with hospitalization data. However, this example has limitations in that not everyone who is hospitalized
needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}.

While
there may be limitations in this specific pairing of a resource and forecasted outcome, we emphasize that the important
result from this work is how evaluating forecasts using a fundamentally different kind of score can illuminate
previously hidden aspects of value in forecasts. In this setting, the feature of a forecast that emerges as being
important is that the probabilistic forecasts predict the relative magnitudes between locations accurately, even if the
absolute magnitudes are inaccurate.

In practice, epidemiological forecasts are often directed at a diverse set of end-users. It may be easy for some of
these users to summarize the consequences of how they use a forecast with a numerical loss function, but for others,
such as officials deciding how best to update public ``situational awareness'' of an outbreak, this may be essentially
impossible. Even those forecast uses that are easily framed as expected loss minimization may differ enough that no
single score would be appropriate for all users. Ideally, targeted forecasting tools could be developed through close
collaboration between modelers and public health officials. However, this may only be possible in settings with
sufficient staffing on both an analytics and a public health team. Increasingly, collaborative modeling hubs are being
used to generate ``one-size-fits-all'' forecasts for many locations at once. But it could still be beneficial in these
settings to expand the set of scoring methods used by the hub to include scores, such as the AS, which are based on
specific public health decision problems.  This could help end users better understand the value of available forecasts
as inputs to their particular decision making contexts. (We note that issues of propriety are raised when forecasts are
evaluated with scoring rules for which they were not elicited or via parameters imputed by a hub after submission, both
of which are done in our application example.  We consider these issues briefly in supplement sections
\ref{sec:distfromq} and \ref{sec:propriety_of_parametric_approximation}.)

### 2024-02-17 alloscore_manuscript/alloscore-application-shorter.tex
attempting to rework/excise ventilator narrative

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Fri Feb 16 13:15:57 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  per-capita & 865 & - & 1029 & 366 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{For one illustrative week, a comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS), (see Section \ref{sec:ias_examp}). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Fri Feb 16 13:16:01 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
model & MAS & taMWIS & taMWIS rank \\ 
  \hline
COVIDhub-ensemble & 389 & 70 & 2 \\ 
  per-capita & 464 & - & - \\ 
  COVIDhub-trained\_ensemble & 483 & 87 & 4 \\ 
  CU-select & 502 & 81 & 3 \\ 
  JHUAPL-SLPHospEns & 526 & 67 & 1 \\ 
  COVIDhub-baseline & 594 & 93 & 5 \\ 
  JHUAPL-Bucky & 643 & 112 & 7 \\ 
  MUNI-ARIMA & 707 & 116 & 8 \\ 
  JHUAPL-Gecko & 929 & 110 & 6 \\ 
  USC-SI\_kJalpha & 1473 & 155 & 9 \\ 
   \hline
\end{tabular}
\caption{Time averaged mean weighted interval scores (taMWIS) and mean allocation scores (MAS) by model across 13 weeks. These results show the average performance across time for the nine models that submitted forecasts for every week from 2021-11-29 through 2022-02-21, as well
  as for a per-capita allocation. MWIS is not defined for the per-capita allocation which does not use forecasts.} 
\label{tab:multi-week-performance-summary}
\end{table}

@@

While our example used a hypothetical scenario to illustrate a potential application of the allocation score, resource
allocation decisions are a realistic example of the kinds of decisions that could motivate more targeted forecasting
exercises.
%One real-world example, which was the motivating example presented in the work above, is allocation of ventilators during a respiratory viral pandemic\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a
scarce resource during outbreaks \citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with
logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not
perfectly so) with hospitalization data. However, this example has limitations in that not everyone who is hospitalized
needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}. While
there may be limitations in this specific pairing of a resource and forecasted outcome, we emphasize that the important
result from this work is how evaluating forecasts using a fundamentally different kind of score can illuminate
previously hidden aspects of value in forecasts. In this setting, the feature of a forecast that emerges as being
important is that the probabilistic forecasts predict the relative magnitudes between locations accurately, even if the
absolute magnitudes are inaccurate.

In practice, there are many potential users of forecasts with many different decision making problems. While outcomes of
some decisions could potentially be specified by loss functions, other uses for forecasts may be less quantifiable, such
as when forecasts are used to inform a general sense of situational awareness that may feed in unmeasurable ways into
other decisions. Even those that are easily framed as expected loss minimization may differ enough that no single score
would be appropriate for all users. Ideally, targeted forecasting tools could be developed through close collaboration
between modelers and public health officials. However, this may only be possible in settings with sufficient staffing on
both an analytics and a public health team. Increasingly, collaborative modeling hubs are being used to generate
``one-size-fits-all'' forecasts for many locations at once. But it could still be beneficial in these settings to expand
the set of scoring methods used by the hub to include scores based on specific public health decision problems, such as
AS, allowing end users to better understand the value of available forecasts as inputs to their particular
decision making contexts. (We note that issues of propriety are raised when forecasts are evaluated with scoring rules
for which they were not elicited or via parameters imputed by a hub after submission, both of which are done in our
application example.  We consider these issues briefly in supplement sections \ref{sec:distfromq} and
\ref{sec:propriety_of_parametric_approximation}.)

>>>>

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Fri Feb 16 20:39:10 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  per-capita & 865 & - & 1029 & 366 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{For one illustrative week, a comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS), (see Section \ref{sec:ias_examp}). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

% latex table generated in R 4.3.1 by xtable 1.8-4 package
% Fri Feb 16 20:39:15 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
model & MAS & taMWIS & taMWIS rank \\ 
  \hline
COVIDhub-ensemble & 389 & 70 & 2 \\ 
  per-capita & 464 & - & - \\ 
  COVIDhub-trained\_ensemble & 483 & 87 & 4 \\ 
  CU-select & 502 & 81 & 3 \\ 
  JHUAPL-SLPHospEns & 526 & 67 & 1 \\ 
  COVIDhub-baseline & 594 & 93 & 5 \\ 
  JHUAPL-Bucky & 643 & 112 & 7 \\ 
  MUNI-ARIMA & 707 & 116 & 8 \\ 
  JHUAPL-Gecko & 929 & 110 & 6 \\ 
  USC-SI\_kJalpha & 1473 & 155 & 9 \\ 
   \hline
\end{tabular}
\caption{Time averaged mean weighted interval scores (taMWIS) and mean allocation scores (MAS) by model across 13 weeks. These results show the average performance across time for the nine models that submitted forecasts for every week from 2021-11-29 through 2022-02-21, as well
  as for a per-capita allocation. MWIS is not defined for the per-capita allocation which does not use forecasts.} 
\label{tab:multi-week-performance-summary}
\end{table}

@@

While our example used a hypothetical scenario to illustrate a potential application of the allocation score, resource
allocation decisions are a realistic example of the kinds of decisions that could motivate more targeted forecasting
exercises.
%One real-world example, which was the motivating example presented in the work above, is allocation of ventilators during a respiratory viral pandemic\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a
scarce resource during outbreaks \citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with
logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not
perfectly so) with hospitalization data. However, this example has limitations in that not everyone who is hospitalized
needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}.

While
there may be limitations in this specific pairing of a resource and forecasted outcome, we emphasize that the important
result from this work is how evaluating forecasts using a fundamentally different kind of score can illuminate
previously hidden aspects of value in forecasts. In this setting, the feature of a forecast that emerges as being
important is that the probabilistic forecasts predict the relative magnitudes between locations accurately, even if the
absolute magnitudes are inaccurate.

In practice, epidemiological forecasts are often directed at a diverse set of end-users. It may be easy for some of these
users to summarize the consequences of how they use a forecast with a numerical loss function, but for others, such
as officials deciding how best to update public ``situational awareness'' of an outbreak, this may be essentially impossible.
Even those forecast uses that are easily framed as expected loss minimization 
may differ enough 
that no single score
would be appropriate for all users. Ideally, targeted forecasting tools could be developed through close collaboration
between modelers and public health officials. However, this may only be possible in settings with sufficient staffing on
both an analytics and a public health team. Increasingly, collaborative modeling hubs are being used to generate
``one-size-fits-all'' forecasts for many locations at once. But it could still be beneficial in these settings to expand
the set of scoring methods used by the hub to include scores, such as the AS, which are based on specific public health decision
problems.  This could help end users better understand the value of available forecasts as inputs to their particular
decision making contexts. (We note that issues of propriety are raised when forecasts are evaluated with scoring rules
for which they were not elicited or via parameters imputed by a hub after submission, both of which are done in our
application example.  We consider these issues briefly in supplement sections \ref{sec:distfromq} and
\ref{sec:propriety_of_parametric_approximation}.)
